% This page is optional but generally included

\begin{acknowledgments}

  Just days after arriving in Bloomington, my wife and I found ourselves at The
  Tap with a few members of the Indiana University Programming Languages group.
  Here we met Cameron Swords, Mike Vitousek, Jaime Guerrero, Matteo Cimini, and
  Jason Hemann, for what would be the first of many long nights spent discussing
  programming languages, science fiction, bad movies, philosophy, and beer.
  Without these nights, I would likely never have survived graduate school.

  My advisor, Ryan Newton, not only taught me how to do research, he gave me a
  framework for how to think about software. He turned me from a Schemer into a
  hardened Haskeller and staunch advocate for static types. He taught me how to
  build a compiler, and building compilers is now one of my favorite things in the
  world.

  I want to thank the other members of my committee, Jeremy Siek, Sam
  Tobin-Hochstadt, and Larry Moss, for everything they have done to help me over
  the years. Similarly, I want to thank Milind Kulkarni for his mentorship
  throughout the last few years.
  I also could not have gotten here without Chaitanya
  Koparkar and Laith Sakka, for their help and collaboration on the Gibbon project.

  For encouraging me to go to graduate school, and for sparking my interest in
  computer science research, I want to thank my instructors at California State
  University, Sacramento: V. Scott Gordon, for showing me Scheme for the first
  time, and for taking the time to talk to me about my personal projects and guiding
  me through the grad school application process; Ted Krovetz, for giving me the
  opportunity to give lectures in his compilers course while I was still an
  undergraduate; and John Clevenger, for giving me the eternally useful advice,
  that ``you don't need to debug your code if you just write it correctly the first time.''

  Early in my Ph.D., I was lucky enough to get to work with Bo Joel Svensson and
  Trevor L. McDonell while they were postdocs at IU, and I want to thank them for
  the mentorship they both provided.

  I am thankful to all the attendees of PL Wonks in Bloomington, for presenting
  their work and putting up with mine. Thanks especially to Wonks regulars Ryan
  Scott, Vikraman Choudhury, Matthew Heimerdinger, Sarah Spall, Laurel Carter,
  Praveen Narayanan, Andre Kuhlenschmidt, Rajan Walia, Buddhika Chamith, Andrew
  Kent, and Paulette Koronkevich. And thanks to Ambrose Bonnaire-Sergeant, for the
  fun conversations and for getting lost with me in Mumbai.

  Thanks to my family for supporting me, for encouraging me when I was down, for
  always telling me they're proud of me, and for politely listening when I tried
  to explain my research.

  And most of all, I want to thank my wife, Victoria Vollmer. She was the person
  who first encouraged me to go to graduate school. At the start, the plan was
  that I would study Computer Science and she would study Mathematics. We applied
  to schools around the country, and eventually, both chose Indiana University. Our
  adventure started with a cross-country drive from California to Indiana, and
  eventually, years later, led to us getting married. Along the way, Victoria had
  become a formidable computer scientist and logician. Thank you, Tori, for being
  my biggest advocate and my best friend.

% This is the (optional) acknowledgments page, which is designed to recognize people or agencies to whom you feel grateful for any academic, technical, financial, or personal aid in the preparation of your thesis. As a matter of courtesy, you should ordinarily mention the members of your committee here, as well as institutions that provided funding.
\end{acknowledgments}
